---
title: "Adventures in Polyglot Packaging"
subtitle: "Stringing together R, C++, WebAssembly, and Javacript with Rcpp, htmlwidgets, and emscripten"
author: "Noam Ross"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
h <- here::here
```

# Introduction 

I've recently become intrigued by [WebAssembly](http://webassembly.org/), a new
technology recently adopted by all the major web browsers for fast program
execution. WebAssembly is essentially a binary set of instructions that can be generated by compiling programs in languages like C++, and then run like JavaScript.  Because the WebAssembly format is compact, does not need to be parsed, and is designed for optimal compiling n browsers, it loads and runs very fast, at speeds approaching native compiled programs.[^wa]

[^wa]: I am sure I am getting some of this wrong.  Lin Clark has an excellent series of
[cartoon explanations of WebAssembly](https://hacks.mozilla.org/2017/02/a-cartoon-intro-to-webassembly/).

Why might WebAssembly be relevant for the R ecosystem?  R already has a lot
of interaction with code in the browser via
[**htmlwidgets**](http://www.htmlwidgets.org/). htmlwidgets execute JavaScript
generated by R functions in the browser, either in pure HTML documents or
in Shiny apps where there is two-way communication between JavaScript and R.
WebAssembly could make the browser-side of this execution faster.

What I find most interesting, though is that WebAssembly can be compiled from
the same C or C++ code that is compiled and run through R directly[^cmp].  That means
that we can write an algorithm once, compile it into both native and WebAssembly
instructions, and then run it in either R or in the browser.

[^cmp]: Technically you could do this with [asm.js](http://asmjs.org/) as well before
WebAssembly came along.

Why would you want to do this?  Well, one might want to use the same algorithm
as part of R-focused model-building and fitting, and then again in _simulating_
or _predicting_ from that already-fit model in the browser.  The former task may be very compute
intensive and require an R development environment and a lot of computer power, 
but the latter task may be incorporated into a user interface in the browser
and require much less power, though still benefit from efficient execution.  htmlwidgets
can link these two tasks, generating browser-based programs based on parameters
passed from R.

You see this paradigm in things like [kerasjs](https://transcranial.github.io/keras-js/#/),
which can make predictions from [Keras](https://keras.io/) models in the browser with no server back-end.
_Fitting_ the Keras models, though, requires the full machinery of Keras and
modeling interfaces like the [R Keras package](https://keras.rstudio.com/) .

In my work, I am interested in stochastic simulation models for epidemic processes such as disease outbreaks.
These simulations can run very fast individually,  but most useful outputs require many repeated simulations.  Fitting them to data is a complex, compute-intensive task for which I generally use [iterated filtering](https://kingaa.github.io/pomp/) methods in R and C++.  A fit model could be usefully deployed as a web page, though, so that practitioners could see simulations, and explore parameter sensitivities or intervention scenarios. 

[This repo](https://github.com/noamross/wassa) is a proof-of-concept exercise in
running an algorithm written in C++ in both R and the browser.  Below I go
through the steps I took to build a minimal package, and then demonstrate the
outputs and performance.

# Writing the core C++ simulator

First, I wrote a standalone C++ simulation function.  All this does is simulate
a Poisson process: counting random, discrete events in a fixed time period.
It's effectively just a random number generator, but I picked it because it's
the simplest component of my ultimate goal, a general [Gillespie algorithm](https://en.wikipedia.org/wiki/Gillespie_algorithm) simulator and its various
accelerated flavors.  This goes in `src/ssa.h`

```{r cppssh, echo=FALSE, comment=""}
cat(readLines(h('src/ssa.h')), sep="\n")
```

Note that, since this code will not only run under R, I'm using C++ random
number generator rather than Rcpp's convenient interface to the ones in the 
R library.  In fact, there's no R or Rcpp-specific code or headers here at all.

# Making wrappers for both R and JavaScript

Next, I wrote a tiny wrapper for this function so as to export it to R via Rcpp
as `rssa()`, and put it in `src/ssa.cpp`.

```{r rcppssh, echo=FALSE, comment=""}
cat(readLines(h('src/ssa.cpp')), sep="\n")
```

Then I wrote _another_ wrapper, this one to export the same function to
WebAssembly and JavaScript via [emscripten](https://kripken.github.io/emscripten-site/), which is the dominant compiler for this.  As this code will eventually be part of an htmlwidget, I put it in.
`inst/htmlwidgets/lib/emssa/emssa.cpp`:

```{r emssh, echo=FALSE, comment=""}
cat(readLines(h('inst/htmlwidgets/lib/emssa/emssa.cpp')), sep="\n")
```

The `EMPSCIPTEN_BINDINGS()` function here takes the `ssa` function defined in `src/ssa.h` and tells empscripten to make it visible in the compiled WebAssembly/JS module it creates.

# Compiling C++ to WebAssembly

I ran emscripten's `em++` compiler to compile the C++ to a WebAssembly module that can be loaded in the browser.  Rather that figure out how to get
emscripten installed, I just used a docker image to run it.  I put this command in `configure`, which probably isn't CRAN-friendly but conveniently runs it before building the package:

```{r emcompile, echo=FALSE, comment=""}
cat(readLines(h('configure')), sep="\n")
```

Some notes on this beast of a command:

-   The first line runs `em++` from the docker image, mounting the project directory into the Docker container's `src/` directory, where compilation happens.
-   The `--bind` argument tells emscripten to use set up bindings to the functions exported in `emssa.cpp`
-  `-s WASM=1` compiles to WebAssembly, and `s SINGLE_FILE=1` tells empscripten to inline the compiled WebAssembly (`.wasm`) file into a `.js` JavaScript wrapper as a data URI.[^1]
- `-O3 ` and `-std=c++11` are C++ compilation settings.
- `-s "EXTRA_EXPORTED_RUNTIME_METHODS=['addOnPostRun']"` gives us this extra method to use in JavaScript code that calls the module.  You'll see what it's for when I build the htmlwidget.

[^1]:  Without `SINGLE_FILE`, emscripten will create  a binary WebAssembly `.wasm` file and a separate JavaScript wrapper  to load it.  This is actually the faster way to go.  `.wasm` binaries load fast and can be compiled while streaming.  However, R's htmlwidgets inline all assets as base64-encoded data URIs anyway, and don't do well with stand-alone binary attachments. In the end, though, this should on affect load time, not run time speed of our function.

After compilation, there's an `emssa.js` file in `inst/htmlwidgets/lib/emssa`.
If you inspect it, you'll see that `wasmBinaryFile` is defined in it as a data URI.

# Making the htmlwidget

As the very handy [htmlwidgets intro](http://www.htmlwidgets.org/develop_intro.html) explains,
I need three files to turn a JavaScript library into an R htmlwidget: the 
dependencies file, the R binding, and the JavaScript binding.  Conveniently,
`htmlwidgets::scaffoldWidget("wassa")` sets up template files for me.  First,
I edited the dependencies file, `inst/htmlwidgets/wassa.yaml`,  to depend on the newly created `emssa.js` file:

```{r depends, echo=FALSE, comment=""}
cat(readLines(h('inst/htmlwidgets/wassa.yaml')), sep="\n")
```
Next, I created the R binding in `R/wassa.R`. This is a function that creates the htmlwidget.
It takes the same arguments as the R and C++ simulator functions,
and passes them to the JavaScript widget as the parameter list `x`.[^2]  The `sizingPolicy` part
just affects how the widget is displayed.
 
 [^2]: `scaffoldWidget()` also creates some Shiny-specific things that I'm dropping for now.


```{r rbinding, echo=FALSE, comment=""}
cat(readLines(h('R/wassa.R')), sep="\n")
```

Finally, I created the JavaScript binding in `inst/htmlwidgets/wassa.js`.  My JS skills are atrocious, but
eventually I got this together:

```{r jsbinding, echo=FALSE, comment=""}
cat(readLines(h('inst/htmlwidgets/wassa.js')), sep="\n")
```

This is based on the `scaffoldWidget()` template. The stuff under `factory:` generates the htmlwidget itself.  When the widget runs the `emssa.js` dependency, it export a `Module` object that contains the compiled functions as well as some helpers. `Module.addOnPostRun`, which I chose to export in the compile step, tells the browser to wait until this the module fully loaded and compiled to run the code. Once that happens, I run the exported compiled function `Module.ssa`, using the parameters we passed the widget from R in `x` variable. I wrap this function in some commands to measure run time, then print the output and timing to the window inside a blue box.

# Building and running

With all these pieces in place, along with a few other essential package bits generated by various **usethis** and **devtools** functions,  I ran `devtools::document()` and `devtools::install()`, and now we can see the results.

First, here's the `rssa()` function which is the Rcpp wrapper for the C++ code:

```{r}
library(wassa)
rssa(1000, 1)
```

Now, htmlwidget version.  If everything works right, you'll see the widget output
in a blue box[^br]:

```{r}
wassa(1000, 1)
```

It works (for me, at least)! My C++ code is running in the browser.  If you refresh the page,
you'll see that the widget output in the blue box changes each time.  No server
is needed - these calculations take place on your machine.

To make sure that both the R and widget versions of the code are doing the 
same thing, I can run the functions with the same random seed and check that outputs
are the same:

```{r}
rssa(1000, 1, seed=10)
wassa(1000, 1, seed=10)
```

ðŸ‘ !

How does performance compare? I run the function at a higher rate, which
will require more iterations, and benchmark:

```{r}
library(microbenchmark)
microbenchmark(rssa(1000000, 1), times = 200, unit = 'ms')
wassa(1000000, 1)
```


Not bad! If you refresh a few times, you'll see the WebAssembly version runs
at 130%-150% of the compiled R native version.  At least, that's the case on _my_ laptop.
What _you_ see is a comparison between the native compiled code on my machine, and
the WebAssembly running on your machine, via the engine in your particular browser.
Note that mileage may vary for native speed, too, not just by machine but also by
the compiler setting used to build the R package.  In my case, the C++ code was compiled
with clang 9.0.0 and he `-O3` flag, the max level of optimization, just as empscripten was.

[^br]:  You need an up-to-date browser to run WebAssembly, but the four major browsers
all had it running as of the end of Feb 2017.  The widget should also work in the latest
RStudio Viewer pane, though sometimes it fails for reasons I haven't quite figured out.
